---
title: "N741 Homework 8"
author: "Julianne"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  pdf_document: default
  html_document: default
  github_document: default
  word_document: default
---

Rcode available here: [https://github.com/JulianneA/N741hmwk8](https://github.com/JulianneA/N741hmwk8)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r, echo=FALSE}
# from tidyverse - use readr
# to read in the comma delimited dataset
library(readr)

# raw data does not have column names
bcdat <- read_csv("breast-cancer-wisconsin.data",
                  col_names=FALSE)

# add variable names
names(bcdat) <- c("idnum","clumpthickness","uniformcellsize",
                  "uniformcellshape","marginaladhesion",
                  "singlecellsize","barenuclei","blandchromatin",
                  "normalnucleoli","mitoses","class")

# note in column 7 "Bare Nucleoli" there are
# question marks "?" that need to be set to missing NA
library(dplyr)
bcdat <- bcdat %>%
  mutate(barenucfix = ifelse(barenuclei=="?",NA,
                              as.numeric(barenuclei)))

# keep the main 11 variables
bcdat <- bcdat %>%
  select(idnum,clumpthickness,uniformcellsize,uniformcellshape,
         marginaladhesion,singlecellsize,barenucfix,blandchromatin,  
         normalnucleoli,mitoses,class)

# keep only complete cases, n=683
bcdat <- na.omit(bcdat)
```
# PCA with full dataset
## 1. Perform the PCA

```{r, echo=FALSE}
# use only columns 2 through 10
# you do not need the idnum, nor the class variables
pr.out <- prcomp(bcdat[,2:10], scale=TRUE)
summary(pr.out)
```

## 2. Make plots of the variance and PVE 

### Plot of the Variances of Each PC

```{r, echo=FALSE}
plot(pr.out)
```

### Plot of the PVE and Cumulative PVE of each PC

```{r, echo=FALSE}
# plots of the PVE percent variance explained
pve = 100*pr.out$sdev^2/sum(pr.out$sdev^2)
pve
plot(pve, type = "o", ylab = "Cumulative PVE", xlab = "Principal Component", col="blue")
plot(cumsum(pve), type = "o", ylab = "Cumulative PVE", xlab = "Principal Component", col="brown3")

```

## 3. Make a "loadings plot" of the variables

```{r, echo=FALSE}
# loadings are in the "rotation" part of the 
# pr.out list object. "rotation" is a matrix
# with a row for each variable and a column for
# each PC.
pr.out$rotation

# choose the 1st and 2nd columns for the 1st 2 PCs
# and plot these loading weights for the 9
# variables. I tweaked the limits some
# feel free to change these as needed
plot(pr.out$rotation[,1],pr.out$rotation[,2],
     xlim=c(-0.5,0.1),ylim=c(-0.5,1),
     cex=2, pch=19,
     xlab = "Principal Component 1",
     ylab = "Principal Component 2",
     main = "Loadings Plot for PC 1 and 2")

# add xpd=FALSE to prevent lines drawn outside plot area
par(xpd=FALSE)

# add red dashed lines for the axes at y=0 and x=0
abline(h=0, col="red")
abline(v=0, col="red")

# overlay the variable names on this loading plot
text(pr.out$rotation[,1],pr.out$rotation[,2],
     labels = rownames(pr.out$rotation),
     pos = 3)
```

## 4. Scores Plot on 1st 2 PCs

```{r, echo=FALSE}
# scores plot - use x from the pr.out list object
# plot scores on 1st 2 PCs, columns 1 and 2 of x
# color the points by the "class" variable for
# benign (class=2) or malignant (class=4)
plot(pr.out$x[,1],pr.out$x[,2], 
     col = bcdat$class,
     xlab = "Principal Component 1",
     ylab = "Principal Component 2",
     main = "Scores Plot on PC 1 and 2",
     sub = "Blue=Benign (class=2) and Red=Malignant (class=4)")
```

# Homework 8 Tasks

```{r, echo=FALSE}
# Benign cases
bcdatBenign <- bcdat %>%
  filter(class == 2)

# Malignant cases
bcdatMalignant <- bcdat %>%
  filter(class == 4)
```

# PCA with Benign Cases only
```{r, echo=FALSE}
# redo for benign ==============
bcdat <- bcdatBenign
# run steps above

```

## 1. Perform the PCA

```{r, echo=FALSE}
# use only columns 2 through 10
# you do not need the idnum, nor the class variables
pr.out <- prcomp(bcdat[,2:10], scale=TRUE)
summary(pr.out)
```

## 2. Make plots of the variance and PVE 

### Plot of the Variances of Each PC

```{r, echo=FALSE}
plot(pr.out)
```

### Plot of the PVE and Cumulative PVE of each PC

```{r, echo=FALSE}
# plots of the PVE percent variance explained
pve = 100*pr.out$sdev^2/sum(pr.out$sdev^2)
pve
plot(pve, type = "o", ylab = "Cumulative PVE", xlab = "Principal Component", col="blue")
plot(cumsum(pve), type = "o", ylab = "Cumulative PVE", xlab = "Principal Component", col="brown3")

```

## 3. Make a "loadings plot" of the variables

```{r, echo=FALSE}
# loadings are in the "rotation" part of the 
# pr.out list object. "rotation" is a matrix
# with a row for each variable and a column for
# each PC.
pr.out$rotation

# choose the 1st and 2nd columns for the 1st 2 PCs
# and plot these loading weights for the 9
# variables. I tweaked the limits some
# feel free to change these as needed
plot(pr.out$rotation[,1],pr.out$rotation[,2],
     xlim=c(-0.5,0.1),ylim=c(-0.5,1),
     cex=2, pch=19,
     xlab = "Principal Component 1",
     ylab = "Principal Component 2",
     main = "Loadings Plot for PC 1 and 2")

# add xpd=FALSE to prevent lines drawn outside plot area
par(xpd=FALSE)

# add red dashed lines for the axes at y=0 and x=0
abline(h=0, col="red")
abline(v=0, col="red")

# overlay the variable names on this loading plot
text(pr.out$rotation[,1],pr.out$rotation[,2],
     labels = rownames(pr.out$rotation),
     pos = 3)
```

## 4. Scores Plot on 1st 2 PCs

```{r , echo=FALSE}
# scores plot - use x from the pr.out list object
# plot scores on 1st 2 PCs, columns 1 and 2 of x
# color the points by the "class" variable for
# benign (class=2) or malignant (class=4)
plot(pr.out$x[,1],pr.out$x[,2], 
     col = bcdat$class,
     xlab = "Principal Component 1",
     ylab = "Principal Component 2",
     main = "Scores Plot on PC 1 and 2",
     sub = "Blue=Benign (class=2)")
```
# PCA with Malignant Cases only
```{r, echo=FALSE}

# redo for malignant ==================
bcdat <- bcdatMalignant
# run steps above
```

## 1. Perform the PCA

```{r, echo=FALSE}
# use only columns 2 through 10
# you do not need the idnum, nor the class variables
pr.out <- prcomp(bcdat[,2:10], scale=TRUE)
summary(pr.out)
```

## 2. Make plots of the variance and PVE 

### Plot of the Variances of Each PC

```{r, echo=FALSE}
plot(pr.out)
```

### Plot of the PVE and Cumulative PVE of each PC

```{r, echo=FALSE}
# plots of the PVE percent variance explained
pve = 100*pr.out$sdev^2/sum(pr.out$sdev^2)
pve
plot(pve, type = "o", ylab = "Cumulative PVE", xlab = "Principal Component", col="blue")
plot(cumsum(pve), type = "o", ylab = "Cumulative PVE", xlab = "Principal Component", col="brown3")

```

## 3. Make a "loadings plot" of the variables

```{r, echo=FALSE}
# loadings are in the "rotation" part of the 
# pr.out list object. "rotation" is a matrix
# with a row for each variable and a column for
# each PC.
pr.out$rotation

# choose the 1st and 2nd columns for the 1st 2 PCs
# and plot these loading weights for the 9
# variables. I tweaked the limits some
# feel free to change these as needed
plot(pr.out$rotation[,1],pr.out$rotation[,2],
     xlim=c(-0.5,0.1),ylim=c(-0.5,1),
     cex=2, pch=19,
     xlab = "Principal Component 1",
     ylab = "Principal Component 2",
     main = "Loadings Plot for PC 1 and 2")

# add xpd=FALSE to prevent lines drawn outside plot area
par(xpd=FALSE)

# add red dashed lines for the axes at y=0 and x=0
abline(h=0, col="red")
abline(v=0, col="red")

# overlay the variable names on this loading plot
text(pr.out$rotation[,1],pr.out$rotation[,2],
     labels = rownames(pr.out$rotation),
     pos = 3)
```

## 4. Scores Plot on 1st 2 PCs

```{r, echo=FALSE}
# scores plot - use x from the pr.out list object
# plot scores on 1st 2 PCs, columns 1 and 2 of x
# color the points by the "class" variable for
# benign (class=2) or malignant (class=4)
plot(pr.out$x[,1],pr.out$x[,2], 
     col = bcdat$class,
     xlab = "Principal Component 1",
     ylab = "Principal Component 2",
     main = "Scores Plot on PC 1 and 2",
     sub = "Red=Malignant (class=4)")
```

# Questions

2. In the overall dataset, when looking at the loadings plot, which variables cluster together? which variables do not lie with that cluster?

In the overall dataset, Mitoses is clearly the most profound outlier, with single cell size and clump thickness lying a bit to the outskirts of the cluster that contains all other variables: Uniformity of Cell Size, Uniformity of Cell Shape, Marginal Adhesion, Bare Nuclei, Bland Chromatin, Normal Nucleoli.

3. How do the variable clusters seen in the loading plots for the Benign data subset and Malignant subset differ? and how are they similar if at all?

The cluster pattern in the loading plot for the benign data subset looks similar to the clustering pattern from the combined dataset, with mitoses as the most profound outlier, although many of the remaining variables, while clustered, are in different locations in the benign plot compared to the whole dataset plot. The loading plot for the malignant subset appears most different from the other two, with the variables more distributed throughout the loading plot relative to the combined and benign plots.
In the malignant plot, clump thickness and bare nuclei are the most distinct outliers, and are more strongly outliers on the x axis (Principal Component 1) compared to the y axis. Additionally, the malignant plot has certain variables, such as bare nuclei and clump thickness with completely flipped values (negative to positive or vice versa) compared to the benign plot.

4. Is using 2 principal components reasonable for summarizing the variability seen in this Breast Cancer dataset with 9 measurements? Explain your reasoning for (a) the overall dataset, (b) the Benign subset and (c) the Malignant subset

The Overall dataset would require the least number of principal components to explain the most variability in the data, as indicated by the scree plots. These plots indicate that using just two PCs would explain approximately 75% of the variability in the data, whereas just two PCs would explain just under 50% of the variability in both the Benign and Malignant subsets. Therefore, based on scree plots, 2 principal components would likely be reasonable for summarizing variability in just the overall dataset. 
The Benign subset has a 2nd large decrease in cumulative PCE between the 6th-7th variables and the malignant subset has a 2nd (relatively) large decrease between the 4th-5th variables and a 3rd between the 8th-9th variables. These "elbows" should be considered when trying to explain variability in the dataset, but they seem to indicate that more variables are required to adequately explain variability in the subsets compared to the full dataset.


5. While PCA is an unsupervised data analysis method (i.e. no "target" class information is used in the analysis), do you think the 2 PCs extracted do a good job of helping to distinguish Benign cases from Malignant cases (i.e. look back at the overall dataset Scores Plot). Explain your rationale.

The 2 PCs extracted seem to do an adequate job of distinguishing between benign and malignant, but would likely be more useful to rule-in malignancy rather than rule it out. The rationale for this is because the malignancy seems to cluster together much more than benign, and while the PCs identified do an adequate job of isolating malignancies, they seem to be more effective at ruling out benign tumors. This means that using the PCs, individuals can fall into "malignant", "likely not malignant", and "definitely not malignant" categories, and additional testing could be conducted on the "likely not malignant" group to continue to rule in/out malignancy.
